Risk Factors and Treatment of Refractory Anaphylaxis - a systematic review

### 2017-12-06 analysis/paper/paper.Rmd
Changes and template problem fix

---
title: 'Risk Factors and Treatment of Refractory Anaphylaxis - a systematic review'
author:
  - Wojciech Francuzik
  - Sabine Dölle
  - Margitta Worm, MD
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
    fig_caption: yes
    reference_docx: templates/template.docx
bibliography: references.bib
csl: mee.csl
abstract: |
  (maximum 200 words).

@@

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

@@

#library(refractoryANA)
require(magrittr,quietly = T)
# source("../../R/data_handling.R")
devtools::load_all(".",quiet = T)

@@

## Epidemiology

The incidence of anaphylaxis is increasing worldwide [@Koplin2011] and was lately estimated to be between 50 and 112 episodes per 100,000 person-years and an estimated prevalence of 0.3% to 5.1% [@TejedorAlonso2015].  It is widely believed to be an underrecognized and undertreated medical emergency [@Fineman2015]. Anaphylaxis in children is in the majority of cases food dependent and most likely more frequently less severe than episodes in adulthood (Ref). Cases of refractory anaphylaxis are extremely rare. The incidence of refractory anaphylaxis was not investigated earlier and it is estimated (based on our data) to range between 3-5% of all cases of anaphylaxis.

## Risk factors for anaphylaxis

@@

## Management

The management algorithms for anaphylaxis are based on consensus expert recommendations and might be incomplete when it comes to reactions that are not responding to the first line therapy. Until to day the acute treatment of anaphylaxis is uniquely recommended in international guidelines and the mainstay of therapy is the early administration of intramuscular (i.m.) adrenaline [@Lazar2017].
However, an optimal anaphylaxis treatment in a given patient is difficult to study because it is an emergency condition and placebo-controlled studies are for ethical reasons limited. Therefore most of the knowledge regarding the treatment of anaphylaxis is based on expert recommendations which are graded as low-level evidence sources.

@@

<!-- We compared the successful procedures from the case reports to ??? -->

<!-- ## Eligibility criteria -->
<!-- Specify study characteristics (e.g., PICOS, length of follow-up) and report characteristics (e.g., years considered, language, publication status) used as criteria for eligibility, giving rationale. -->

<!-- ## Information sources -->
<!-- Describe all information sources (e.g., databases with dates of coverage, contact with study authors to identify additional studies) in the search and date last searched. -->

## Literature search strategy

@@

We collected information about the elicitors, clinical setting, concomitant diseases, medication, reaction symptoms, therapy and outcomes of the treatment. We used piloted polls to gather the data and all reports were screened once by the same researcher. We investigated which interventions were reported as successful in the treatment of anaphylactic hypotension that was not responding to injected (i.v. or i.m.) adrenaline. Study outcomes were defined as time to elicit a response and mortality. Successful and unsuccessful management of anaphylaxis was one of the most important outcomes we evaluated. This information was available in all cases. All of the studies were case reports therefore the evidence level was classified as low. The review protocol was not established prior to the review and the review did not undergo preregistration.

<!-- ## Data items -->
<!-- List and define all variables for which data were sought (e.g., PICOS, funding sources) and any assumptions and simplifications made. NP GROUPS THAT WERE MADE OUT OF DATA INPUTS!!!! -->

<!-- ## Risk of bias in individual studies -->
<!-- Describe methods used for assessing the risk of bias of individual studies (including specification of whether this was done at the study or outcome level), and how this information is to be used in any data synthesis. -->

<!-- ## Summary measures -->

<!-- State the principal summary measures (e.g., risk ratio, difference in means). -->

<!-- ## Synthesis of results -->

<!-- Describe the methods of handling data and combining results of studies, if done, including measures of consistency (e.g., I2) for each meta-analysis. -->

<!-- ## Risk of bias across studies  -->

<!-- Specify any assessment of risk of bias that may affect the cumulative evidence (e.g., publication bias, selective reporting within studies).  -->

<!-- ## Additional analyses  -->

<!-- Describe methods of additional analyses (e.g., sensitivity or subgroup analyses, meta-regression), if done, indicating which were pre-specified.  -->

```{r demograph}
knitr::kable(demogr.tab,col.names = c("Elicitor","No. cases","Age","Male fraction [%]","Surgical intervention [%]"),digits = 1)
```

@@

A major risk factor for refractory anaphylaxis is perioperative anaphylaxis. In `r (d$Situation_gr=="OP") %>% which %>% length()` (`r ((d$Situation_gr=="OP") %>% which %>% length()/length(d[,1])*100) %>% round(1)`%) cases the anaphylactic reaction occured during a diagnostic or surgical procedure. The most common causes of refractory anaphylaxis based on the analyzed case reports were iatrogenic procedures including anesthesia drugs, aprotinin, and protamine, radiocontrast media (RCM). Surprisingly, no cases of insect venom anaphylaxis were found among these reports. While the incidence of perioperative anaphylaxis has been reported to be between 1 in 10,000 – 20,000 anesthesia procedures, it has a relatively high fatality rate of 3 – 10% of the perioperative fatalities. [@Nag2017]

<!-- # Firs sings in anesthesia setting of anaphylaxis -->
An early diagnosis of anaphylaxis during operative procedures can be suggested based on end-tidal CO~2~ collapse, which often is preceding cardiac arrest, even in the absence of any cutaneous manifestations. [@Baumann2009] In such case - treatment with adrenalin should be immediately initiated.

@@

Most of the treatment strategies were according to the guidelines of anaphylaxis management [@Simons2011] but were supplemented by additional procedures that have saved patients' lives when first-line therapy failed. We report the most frequent procedures in order to evaluate them in the treatment of refractory anaphylaxis.

Based on the analyzed reports, ambulatory patients were first treated with adrenaline were patients experiencing anaphylaxis under anesthesia were usually first given intravenous volume replacement and 100% oxygen, followed by steroid and antihistaminic agents. Epinephrine was given in 100% of cases in various doses.

@@

4 fatal cases (`r (length(which(d$Death=="yes"))/length(d$Death)*100) %>% round(1)`%) which did not respond to any therapy are listed in the table below.

```{r tabFatal}
rownames(fat) <- NULL
colnames(fat) <- c("Elicitor","Age","Sex","Situation","Steroids","Antihistamines","Epinephrine [mg]", "Fluids [L]","Vasopressors"  )
knitr::kable(fat)
```

@@

Clinicians must be aware that methylene blue may provoke false pulse oximetry readings as the blue dye mimics cyanosis [@Kessler1986].

@@

In case of a hypotension refractory to adrenaline or volume replacement therapy -- vasopressors can be evaluated.

Vasopressors were used in `r (d$t_vasopression==T) %>% sum` (`r (((d$t_vasopression==T) %>% sum)/((d$t_vasopression==T) %>% length)*100) %>% round(1)`%) of refractory cases. 
Vasopressin or its synthetic analogue terlipressin were used in `r sum(d$t_vasopressin)` cases. In 8 of these reports (`r (8/sum(d$t_vasopressin)*100) %>% round(1)`%) injection of vasopressin reversed the refractory hypotension within minutes. 
Other vasopressors included metaraminol 
[`r which.sources(d$Source,"metara",list(d$What.helped,d$Therapy))`], methoxamine 
[`r which.sources(d$Source,"methox",list(d$What.helped,d$Therapy))`], dopamine 
[`r which.sources(d$Source,"dopam",list(d$What.helped,d$Therapy))`], and noradrenaline [`which.sources(d$Source,"nor",list(d$What.helped,d$Therapy))`] infusions.
Drugs given perioperatively were the elicing factors in all but two of these cases (`r round((length(d[temp.v,1])-2)/length(d[temp.v,1])*100,1)`%) and all patients reacted within minutes (range 5 - 20) upon exposition to the suspected drug. One patient rected to systemic immune therapy (SIT) injection that was inadverently extremely high-concentrated [@Bickell1984] and in one case the elicitor could not be established [@Heytman2004].

Use of vasopressin in refractory anaphylaxis was only suggested in  management recommendations from the AAAAI [@Lieberman] and was never vastly investigated. Dopamine, however, is recommended by 2 out of three guidelines. Nevertheless, dopamine might be related to higher risk of mortality and adverse events than epinephrine or norepinephrine [@Avni2015]. Both norepinephrine and vasopressin may be appropriate second-line agents for the management of refractory hypotension. Vasopressin acts on the V1 receptors, while Norepinephrine acts primarily on the alfa-adrenergic receptors, their synergistic effects may result in effective vasoconstriction [@DiChiara2008].

@@

One case report described the use of military anti-shock trousers to perform external vasopressin on a patient who received inadvertently high doses of systemic immune therapy allergen [@Bickell1984]. Although this can be hardly available in critical care centers we would like to point out this minimally invasive method as a possible alternative measure in achieving normotension.

The European anaphylaxis register lists 18 cases where dopamine was used. None of these reactions were fatal and `r ((dopam.anareg$d_severity_rm %>% summary())[4] / dopam.anareg$d_severity_rm %>% summary() %>% sum()*100) %>% round(1)`% were Ring and Messmer grade IV. `r ((dopam.anareg$d_elicitor_gr5=="drugs") %>% sum()/length(dopam.anareg$d_elicitor_gr5)*100) %>% round(1)`% of there reactions were elicited by drugs or occured during a medical procedure. Adrenaline was not used in `r ((dopam.anareg$d_elicitor_gr5=="drugs") %>% sum()/length(dopam.anareg$d_elicitor_gr5)*100) %>% round(1)`%

@@

Adrenaline may fail to inhibit an anaphylctic event in patients taking beta-blockers as both heart and lungs possess beta-adrenergic-receptors.

Glucagon, a polypeptide hormone with potent inotropic and chronotropic actions was given in two cases of refractory hypotension in analyzed cases[@Zaloga1986; @Javeed1996]. Both patients recieiving glucagon were undergoing refractory anaphylaxis with the concomitant use of atenolol for arterial hypertension. They both reacted to contrast medium and after minutes they have devloped severe anaphylaxis not responding to repeated doses of adrenaline. Infusion of 1 mg glucagon per hour or 1 mg glucagon as a bolus releaved the refractory shock in both cases and it also sucessfully treated a recurrent biphasic reaction [@Javeed1996].

@@

In the most extreme cases, where all pharmacologic attempts fail to alleviate symptoms of anaphylaxis clinicians turned to last resort extracorporeal life support (ECLS) options. There were 4 reports where patients reacted extremely severe and all were operated on because of a heart condition in general anesthesia [`r which.sources(d$Source,"ECLS",d$What.helped_gr)`]. Two patients reacted to chlorhexidine, one to recuronium and in one case the elicitor could not be established. Only one of these patients was taking a beta blocker [@Yee2016]. All these patients recieved multiple doses of adrenaline 1 -- 15 mg, vasopressors, high volume replacement therapy, coricostreroids and antihistaminics. Upon prolonged resuscitation procedures two of these patients were place on cardiopulmonary bypass (CPB) and two on extracorporeal membrane oxygenation (ECMO) devices. After stabilizing of their condition, patients were weaned out from these supportive measures without neurological sequele.

Life-threatening anaphylactic reactions occur in approximately 1 in 6000 anesthetic procedures and are associated with 5% mortality [@Weiss2015]. ECMO is widely used for the treatment of cardiogenic shock [@Sidebotham2010] and may also be used in cases of anaphylaxis. It is especially helpful where there is refractory anaphylaxis with cardiac or respiratory arrest. Its use may gain time for to get rid of the elicitor (metabolize the drug or find the elicitor). One of the reported cases identified the clorhexidin coated cathether to be responsible for the refractory shock, and after its removal patient responded immidiately to therapy. The use of ECMO facilitated the elicitor discovery [@Wang2016].

# Conclusion 
<!-- **The conclusion for all articles should contain a brief summary of the data presented in the article. Please note that this section is meant to be distinct from, and appear before the ‘Expert opinion’ section.** -->

We analysed case reports of refractory anaphylaxis and evaluated the undertaken therapeutic aproaches. The cases were mostly elicited by drugs and happend during a medical procedure. Patients were in all cases treated with adrenaline as a first line therapy, nearly all of them were given oxygen, steroids i.v. and fluid replacement therapy.

We highlight therapeutic options that should be evaluated in case of anaphylaxis non-responsive to first line therapy with adrenaline: use of mythylene blue, vasopressors, glucagon and extracorporeal life support. We suggest a managment algorithm for refractory anaphylaxis (see Fig. \@ref(fig:algo).

# Expert Commentary: 500-1000 words (included in overall word count).

Severe anaphylaxis is a critical medical condition requiring an immediate intervention. According to the international guidelines adrenaline given intramuscular (i.m.) is a rapid and relatively safe treatment stabilizing the symptoms quickly in a given patient. However, in a few patients this intervention might not be sufficient and, concomitantly to repeating its doses, other therapeutical measures will need to be applied. Such cases are regarded as refractory anaphylaxis, which are particulary seen, when the elicitor of the reaction is reaching the organism intravenously. This route is associated with a high systemic load of a given allergen or other mast cell activating molecule inducing a rapid onset of mediator release like histamine, but also tryptase and chymase, which are proteases and which may in addition to histamine promote severely anaphylaxis via activation of the plasma kallikrein and to enhancment of vascular permeability [@Imamura1996].

@@

* Epinephrine in appropriate doses is still the mainstay of anaphylaxis therapy without any absolute contraindications.  for its use in treating anaphylaxis. ? Delay in administration of epinephrine may lead to more severe and treatment resistant anaphylaxis.*[@Fineman2015]
* Glucagon
* Vasopressors
* ECLS
*

# References
<div id="refs"></div>
<!-- **A maximum of 100 references is suggested. Ensure that all key work relevant to the topic under discussion is cited in the text and listed in the bibliography. Reference to unpublished data should be kept to a minimum and authors must obtain a signed letter of permission from cited persons to use unpublished results or personal communications in the manuscript.** -->

## Annotated bibliography
<!-- **Important references should be highlighted with a one/two star system and brief annotations should be given (see the journal’s Instructions for Authors page for examples and for a more detailed description of our referencing style).** -->

# Figures and Tables
<!-- **Up to 5 figures and 5 tables are permitted. For further information on tables and figures, please see our formatting guide**. -->

```{r prismaflow, echo=F,fig.cap="The PRISMA flowchart for included sources"}
#source("../../R/prisma.R")
knitr::include_graphics("../figures/prism_flow.png")
```

```{r algo, echo=F,fig.cap="The algorithm for refractory anaphylaxis management"}
#source("../../R/prisma.R")
knitr::include_graphics("../figures/algo.png")
```

```{r wholeTab}
knitr::kable(whole.tab)
```

```{r dopamine}
knitr::kable(dopam.anareg[,c(-3,-5)])
```

<!-- ### Colophon -->

<!-- This report was generated on `r Sys.time()` using the following computational environment and dependencies:  -->

```{r colophon, cache = FALSE}
# which R packages and versions?
# devtools::session_info()
```

<!-- The current Git commit details are: -->

```{r}
# what commit is this file at? You may need to change the path value
# if your Rmd is not in analysis/paper/
# git2r::repository("../..")
```

>>>>

---
title: 'Risk Factors and Treatment of Refractory Anaphylaxis - a systematic review'
author:
  - Wojciech Francuzik
  - Sabine Dölle
  - Margitta Worm, MD
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  bookdown::word_document2:
    fig_caption: yes
    reference_docx: "../templates/template.docx"
bibliography: references.bib
csl: mee.csl
abstract: |
  (maximum 200 words).

@@

@@

require(magrittr,quietly = T)
devtools::load_all(".",quiet = T)

@@

The incidence of anaphylaxis is increasing worldwide [@Koplin2011] and was lately estimated to be between 50 and 112 episodes per 100,000 person-years and an estimated prevalence of 0.3% to 5.1% [@TejedorAlonso2015].  It is widely believed to be an underrecognized and undertreated medical emergency [@Fineman2015]. Anaphylaxis in children is in the majority of cases food dependent and most likely less severe than episodes in adulthood [@Braganza2005]. Cases of refractory anaphylaxis are extremely rare. The incidence of refractory anaphylaxis was not investigated earlier and it is estimated (based on our data) to range between 3--5% of all cases of anaphylaxis.

@@

The management algorithms for anaphylaxis are based on consensus expert recommendations and might be incomplete when it comes to reactions that are not responding to the first line therapy. To date, the acute treatment of anaphylaxis is uniquely recommended in international guidelines and the mainstay of therapy is the early administration of intramuscular (i.m.) adrenaline [@Lazar2017].
However, an optimal anaphylaxis treatment in a given patient is difficult to study because it is an emergency condition and placebo-controlled studies are limited for ethical reasons. Therefore most of the knowledge regarding the treatment of anaphylaxis is based on expert recommendations which are graded as low-level evidence sources.

@@

@@

We collected information about the elicitors, clinical setting, concomitant diseases, medication, reaction symptoms, therapy and outcomes of the treatment. We used piloted polls to gather the data and all reports were screened once by the same researcher. We investigated which interventions were reported as successful in the treatment of anaphylactic hypotension that was not responding to injected (i.v. or i.m.) adrenaline. Successful and unsuccessful management of anaphylaxis was one of the most important outcomes we evaluated. This information was available in all cases. All of the studies were case reports therefore the evidence level was classified as low. The review protocol was not established prior to the review and the review did not undergo preregistration.

@@

A major risk factor for refractory anaphylaxis is perioperative anaphylaxis. In `r (d$Situation_gr=="OP") %>% which %>% length()` (`r ((d$Situation_gr=="OP") %>% which %>% length()/length(d[,1])*100) %>% round(1)`%) cases the anaphylactic reaction occured during a diagnostic or surgical procedure. The most common causes of refractory anaphylaxis based on the analyzed case reports were iatrogenic procedures including anesthesia drugs, aprotinin, and protamine, radiocontrast media (RCM). Surprisingly, no cases of insect venom anaphylaxis were found among these reports (see Table \@ref(tab:demograph)). While the incidence of perioperative anaphylaxis has been reported to be between 1 in 10,000 -- 20,000 anesthesia procedures, it has a relatively high fatality rate of 3 -- 10% of the perioperative fatalities. [@Nag2017]

An early diagnosis of anaphylaxis during operative procedures can be suggested based on end-tidal CO~2~ collapse, which often is preceding cardiac arrest, even in the absence of any cutaneous manifestations. [@Baumann2009] In such case - treatment with adrenalin should be immediately initiated as delayed adrenaline use has been linked with increased mortality [@Simons2010] and refractoriness to adrenaline [@Fineman2015].

@@

Most of the treatment strategies were in accordance with the guidelines of anaphylaxis management [@Simons2011] but were supplemented by additional procedures that have saved patients' lives when first-line therapy failed. We report the most frequent procedures in order to evaluate them in the treatment of refractory anaphylaxis.

Based on the analyzed reports, ambulatory patients were first treated with adrenaline, whereas patients experiencing anaphylaxis under anesthesia were usually first given intravenous volume replacement and 100% oxygen, adrenaline followed by steroid and antihistaminic agents. Epinephrine was given in all cases in various doses.

@@

4 fatal cases (`r (length(which(d$Death=="yes"))/length(d$Death)*100) %>% round(1)`%) which did not respond to any therapy are listed in the table \@ref(tab:tabFatal)

@@

Clinicians must be aware that methylene blue may provoke false pulse-oximetry readings as the blue dye mimics cyanosis [@Kessler1986].

@@

In case of a hypotension refractory to adrenaline or volume replacement therapy -- vasopressors can be evaluated. Vasopressors were used in `r (d$t_vasopression==T) %>% sum` (`r (((d$t_vasopression==T) %>% sum)/((d$t_vasopression==T) %>% length)*100) %>% round(1)`%) of refractory cases. 
Vasopressin or its synthetic analogue terlipressin were used in `r sum(d$t_vasopressin)` cases. In 8 of these reports (`r (8/sum(d$t_vasopressin)*100) %>% round(1)`%) injection of vasopressin reversed the refractory hypotension within minutes. 
Other vasopressors included metaraminol 
[`r which.sources(d$Source,"metara",list(d$What.helped,d$Therapy))`], methoxamine 
[`r which.sources(d$Source,"methox",list(d$What.helped,d$Therapy))`], dopamine 
[`r which.sources(d$Source,"dopam",list(d$What.helped,d$Therapy))`], and noradrenaline [`r which.sources(d$Source,"nor",list(d$What.helped,d$Therapy))`] infusions.
Drugs given perioperatively were the elicing factors in all but two of these cases (`r round((length(d[temp.v,1])-2)/length(d[temp.v,1])*100,1)`%) and all patients reacted within minutes (range 5 - 20) upon exposition to the suspected drug. One patient reacted to systemic immune therapy (SIT) injection that was inadverently extremely high-concentrated [@Bickell1984] and in one case the elicitor could not be established [@Heytman2004].

Use of vasopressin in refractory anaphylaxis was only suggested in  management recommendations from the AAAAI [@Lieberman] and was never vastly investigated. Dopamine, however, is recommended by 2 out of three guidelines. Nevertheless, dopamine might be related to higher risk of mortality and adverse events than epinephrine or norepinephrine [@Avni2015]. Both norepinephrine and vasopressin may be appropriate second-line agents for the management of refractory hypotension. Vasopressin acts on the V1 receptors, while norepinephrine acts primarily on the alfa-adrenergic receptors, their synergistic effects may result in effective vasoconstriction [@DiChiara2008].

@@

One case report described the use of military anti-shock trousers to perform external vasopression on a patient who received inadvertently high doses of systemic immune therapy allergen [@Bickell1984]. Although this can be hardly available in critical care centers we would like to point out this minimally invasive method as a possible alternative measure in achieving normotension.

The European anaphylaxis register lists 18 cases where dopamine was used (mean age `r mean(dopam.anareg$d_age) %>% round(1)`; range `r dopam.anareg$d_age %>%  range() %>% paste0(collapse="--")`). None of these reactions were fatal and `r ((dopam.anareg$d_severity_rm %>% summary())[4] / dopam.anareg$d_severity_rm %>% summary() %>% sum()*100) %>% round(1)`% were Ring and Messmer grade IV. `r ((dopam.anareg$d_elicitor_gr5=="drugs") %>% sum()/length(dopam.anareg$d_elicitor_gr5)*100) %>% round(1)`% of there reactions were elicited by drugs or occured during a medical procedure. Adrenaline was not used in `r ((dopam.anareg$d_522_adren_agg=="no") %>% sum()/length(dopam.anareg$d_elicitor_gr5)*100) %>% round(1)`%

@@

Adrenaline may fail to inhibit an anaphylctic event in patients taking beta-blockers as both heart and lungs possess beta-adrenergic receptors.
Glucagon, a polypeptide hormone with potent inotropic and chronotropic actions was given in two cases of refractory hypotension in analyzed cases[@Zaloga1986; @Javeed1996]. Both patients recieiving glucagon were undergoing refractory anaphylaxis with the concomitant use of atenolol for arterial hypertension. They both reacted to contrast medium and after minutes they have devloped severe anaphylaxis not responding to repeated doses of adrenaline. Infusion of 1 mg glucagon per hour or 1 mg glucagon as a bolus releaved the refractory shock in both cases and it also sucessfully treated a recurrent biphasic reaction [@Javeed1996].

@@

In the most extreme cases, where all pharmacologic attempts fail to alleviate symptoms of anaphylaxis clinicians turned to last resort extracorporeal life support (ECLS) options. There were 4 reports where patients reacted extremely severe and all were operated on because of a heart condition in general anesthesia [`r which.sources(d$Source,"ECLS",d$What.helped_gr)`]. Two patients reacted to chlorhexidine, one to recuronium and in one case the elicitor could not be established. Only one of these patients was taking a beta blocker [@Yee2016]. All these patients recieved multiple doses of adrenaline (total dose ranged 1 -- 15 mg), vasopressors, high volume replacement therapy, coricostreroids and antihistaminics. Upon prolonged resuscitation procedures two of these patients were place on cardiopulmonary bypass (CPB) and two on an extracorporeal membrane oxygenation (ECMO) device. After stabilizing of their condition, patients were weaned out from these supportive measures without neurological sequele.

Life-threatening anaphylactic reactions occur in approximately 1 in 6000 anesthetic procedures and are associated with 5% mortality [@Weiss2015]. ECMO is widely used for the treatment of cardiogenic shock [@Sidebotham2010] and may also be used in cases of anaphylaxis. It is especially helpful where there is refractory anaphylaxis with cardiac or respiratory arrest. Its use may gain time for to eliminate the responsible elicitor (metabolize the drug or find the occluded elicitor). One of the reported cases identified the clorhexidin coated cathether to be responsible for the refractory shock, and after its removal patient responded immidiately to therapy. The use of ECMO facilitated the elicitor discovery [@Wang2016].

# Conclusion

We analysed case reports of refractory anaphylaxis and evaluated the undertaken therapeutic approaches. The cases were most frequently elicited by drugs and happend during a medical procedure. All patients were treated with adrenaline as a first line therapy, nearly all of them were given oxygen, steroids i.v. and fluid replacement therapy.

We highlight therapeutic options that should be evaluated in case of anaphylaxis non-responsive to first line therapy with adrenaline: methylene blue, vasopressors, glucagon and extracorporeal life support. We suggest a managment algorithm for refractory anaphylaxis (see Fig. \@ref(fig:algo)) that is supplementary to the established protocols.

# Expert Commentary

Severe anaphylaxis is a critical medical condition requiring an immediate intervention. According to the international guidelines adrenaline given intramuscular (i.m.) is a rapid and relatively safe treatment stabilizing the symptoms quickly in a given patient. However, in a few patients this intervention might not be sufficient and, concomitantly to repeating its doses, other therapeutical measures will need to be applied. Such cases are regarded as refractory anaphylaxis, which are particularly seen, when the elicitor of the reaction is reaching the organism intravenously. This route is associated with a high systemic load of a given allergen or other mast cell activating molecule inducing a rapid onset of mediator release like histamine, but also tryptase and chymase, which are proteases and which may in addition to histamine promote severely anaphylaxis via activation of the plasma kallikrein and to enhancment of vascular permeability [@Imamura1996].

@@

* Epinephrine in appropriate doses is still the mainstay of anaphylaxis therapy without any absolute contraindications and should be given as soon as anaphylaxis is suspected. 
* Glucagon infusion should be considered when patients recieve beta-blockers.
* Methylene blue and Vasopressors should be considered together with volume replacement therapy in cases of refractory hypotension. 
*

# Annotated bibliography
<div id="refs"></div>
<!-- **A maximum of 100 references is suggested. Ensure that all key work relevant to the topic under discussion is cited in the text and listed in the bibliography. Reference to unpublished data should be kept to a minimum and authors must obtain a signed letter of permission from cited persons to use unpublished results or personal communications in the manuscript.** -->

<!-- **Important references should be highlighted with a one/two star system and brief annotations should be given (see the journal’s Instructions for Authors page for examples and for a more detailed description of our referencing style).** -->

# Figures and Tables

```{r prismaflow, echo=F,fig.cap="The PRISMA flowchart for included sources"}
knitr::include_graphics("../figures/prism_flow.png")
```

```{r algo, echo=F,fig.cap="The algorithm for refractory anaphylaxis management"}
knitr::include_graphics("../figures/algo.png")
```

```{r demograph}
knitr::kable(demogr.tab,col.names = c("Elicitor","No. cases","Age","Male fraction [%]","Surgical intervention [%]"),digits = 1,
             caption = "Demography of refractory anaphylaxis cases.")
```

```{r tabFatal}
rownames(fat) <- NULL
colnames(fat) <- c("Elicitor","Age","Sex","Situation","Steroids","Antihistamines","Epinephrine [mg]", "Fluids [L]","Vasopressors"  )
knitr::kable(fat,
             caption = "Fatal anaphylaxis cases, not responsive to any form of therapy.")
```

```{r wholeTab}
knitr::kable(whole.tab,
             caption = "List of papers and anaphylaxis cases included in the sudy, summarizing their elicitors, effective interventions and therapy.")
```
